Algorithm explanation

### 2013-05-13 article/algo.tex
Changes of Horn graph structure

The Horn clause solving algorithm receives an input problem in Horn
graph. A \textbf{Horn graph} is a labelled directed graph $G=(V,E,\varphi)$.
\begin{itemize}
\item $V$ is a union of two sets of vertices; Horn term vertices $V_T$
  and arrow vertices $V_\Rightarrow$. Each vertex $u \in
  V_\Rightarrow$ has eactly one outgoing edge to some $v \in V_T$,
  denoted as $succ(u)$.
\item $E$ is a set of labelled edges, which are expressed in the form
  $(u,v,\theta)$ where $\theta$ is a finite map from variables to
  variables.
\item $\varphi: V_T \rightarrow L$ is a labelling function for Horn
  term vertices, where L is a set of elements of a predicate variable
  in the form $P(x_1, ..., x_{arity(P)})$ or a linear inequality $a_1
  x_1 + \cdots + a_n x_n \leq b$.
\end{itemize}

There are two kinds of edges in a Horn graph;
\begin{itemize}
\item for $u \in V_T, v \in V_\Rightarrow$, an edge takes a form
  $(u,v,\theta)$, or,
\item for $u \in V_\Rightarrow, v \in V_T$, an edge takes a form
  $(u,v,\emptyset)$, where the mapping is empty.
\end{itemize}

The meaning $\llbracket G \rrbracket $ of a Horn graph $G$ is a set of
Horn clauses $\mathcal{HC}_v$ for all vertices $v \in V_\Rightarrow$.

\begin{align*}
\mathcal{HC}_v & = \left( \bigwedge_{(u,v,\theta) \in E} \theta \varphi(u) \right) \Longrightarrow \varphi(succ(v)) \\
\llbracket G \rrbracket & = \bigwedge_{v \in V_\Rightarrow} \mathcal{HC}_v
\end{align*}

We implicitly assume each $\mathcal{HC}_v$ is universally quantified
by its all free variables $\textsc{FV}(\mathcal{HC}_v)$. Without loss
of generality, we limit occurrence of linear expression on right-hand
side of Horn clauses to either $0 \leq -1$, meaning $\bot$, or $0 \leq
0$, meaning $\top$.

Note that we allow multiple appearance of Horn clauses with the same
predicate variable on their right-hand side.

@@

We consider below a Horn graph that is acyclic.

For simplicity of discussion, we define $k_v$ the arity of the
right-hand side of $\mathcal{HC}_v$. Assuming that variables in every
Horn clauses are appropriately renamed, the right-hand side of
$\mathcal{HC}_v$ takes the form $P(x_1,...,x_{k_v})$ where $k_v =
arity(P)$, or either $\top$ or $\bot$ with $k_v = 0$. Also we let $K$
denote the maximum of $k_v$ over all $v$.

Given an acyclic Horn graph (or called a Horn DAG), we define the
relation $u \xrightarrow{\theta} v$ over $V_T$ as follows.
\[ u \xrightarrow{\theta} v \Longleftrightarrow
\exists a \in V_\Rightarrow; \left\lbrace (u,a,\theta), (a,v,\emptyset) \right\rbrace \subseteq E \]
We may omit $\theta$ and write $u \rightarrow v$ for convenience for
later discussion. By the assumption that $G$ is acyclic,
$\rightarrow^+$ is a well-founded relation. We assume without loss of
generality that $V_T$ contains the greatest element $v_\bot$ with
respect to $\rightarrow^+$; otherwise we can always implicitly assume
$\varphi(v_\bot) = \top$ such that $\forall v \in V_T; v
\xrightarrow{\emptyset} v_\bot$.

For each $v \in V_T$, we define
$\Delta(v) = (\mathbf{a}_v \mathbf{x}_v \leq b_v, C_v)$
, where
\begin{itemize}
\item $\mathbf{a}_v \mathbf{x}_v \leq b_v$ is a linear inequality that
  contains coefficient variables $\mathbf{a}_v$, and
\item $C_v$ is a set of linear expressions on coefficient variables.
\end{itemize}
Intuitively, the pair $(\mathbf{a} \mathbf{x} \leq b, C)$ denotes the
set of inequalities:

\begin{align*}
\left\lbrace
 \mathbf{a} \mathbf{x} \leq b \middle|
 \exists \left( \textsc{FV}(C)
  \setminus (\mathbf{a} \cup \left\lbrace b \right\rbrace
 \right); C
\right\rbrace
\end{align*}

$\Delta(v)$ is defined by well-founded induction on $v$ with respect
to the relation $\rightarrow^+$. Then, $C_v$ in
$\Delta(v) = \left( \mathbf{a}_v \mathbf{x}_v \leq b_v, C_v \right)$
can be shown as follows.

\begin{align*}
C_v = & \bigcup_{u \rightarrow v} C_u
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u \rightarrow v \\
\left\lbrace
 \mathbf{a}_{v,i} = \sum_{u \xrightarrow{\theta} v} \mathbf{a}_{u, \theta^{-1} (i)},
 b_v \geq \sum_{u \xrightarrow{\theta} v} b_u
\right\rbrace \cup \\
\hspace{2cm} \bigcup_{k_v < i \leq K} \left\lbrace \mathbf{a}_{v,i} = 0 \right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}_v) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_v = \lambda_v \mathbf{a}_0,
 b_v = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x}_v \leq b_0
\end{cases}
\end{align*}

@@

@@

We provide $\textsc{Copy}_v$ procedure to rename all free variables
for avoiding name collision.
\[ \textsc{Copy}_v (x_V) = x_{V \cup \left\lbrace v \right\rbrace} \]
All variables are then distinguished by the label of copy origin
nodes. For consistency, the variables without the label should be
interpreted as that it has $\emptyset$ for origin nodes.

We give a constraint function over a copy tree
$\hat \Delta(v,e)$ for $v \in V_C, e \in E$ as
\[ \hat \Delta (v,e) = \textsc{Copy}_v (\Delta_{V'} (\xi (v)))
\text{ such that } V' = \left\lbrace v' \middle| (v, v',e) \in E_S \right\rbrace \]
extending $\Delta$ function to
$\Delta_V (v) = (\mathbf{a}_{v,\emptyset} \mathbf{x}_v \leq b_{v,\emptyset}, C_v)$
as
\begin{align*}
C_v = & \bigcup_\circ C_u \cup \bigcup_\bullet C_u \cup
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u \rightarrow v \\
\left\lbrace
 \mathbf{a}_{v,\emptyset,i} =
  \sum_\circ \mathbf{a}_{u,U,\theta^{-1}(i)} +
  \sum_\bullet \mathbf{a}_{u,U, \theta^{-1}(i)}
\right\rbrace \cup \\
\hspace{2cm} \left\lbrace
 b_{v,\emptyset} \geq
  \sum_\circ b_{u,U} +
  \sum_\bullet b_{u,U}
\right\rbrace \cup \\
\hspace{2cm} \bigcup_{k_v < i \leq K} \left\lbrace \mathbf{a}_{v,\emptyset,i} = 0 \right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}_v) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_{v,\emptyset} = \lambda_v \mathbf{a}_0,
 b_{v,\emptyset} = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x}_v \leq b_0
\end{cases}
\end{align*}
under
\[
\circ = \substack{
  u \rightarrow v \\
  \nexists u' \in V; \xi (u') = u \\
  \Delta_V (u) = (\mathbf{a}_{u,U} \mathbf{x}_u \leq b_{u,U}, C_u)}
\qquad \bullet = \substack{
  u \rightarrow v \\
  \exists u' \in V; \xi (u') = u \\
  \hat \Delta (u', u \rightarrow v) = (\mathbf{a}_{u,U} \mathbf{x}_u \leq b_{u,U}, C_u)} \]

@@

\begin{align*}
 \rho = \left\lbrace
  \left( P, \bigwedge_S \sigma(\mathbf{a}_{v,S} \mathbf{x}_v \leq b_{v,S}) \right) \middle|
  \forall v \in V_T; \varphi(v) = P(\mathbf{x}_v)
 \right\rbrace
\end{align*}

@@

>>>>

\subsection{Preliminaries}

The Horn clause solving algorithm receives an input problem in Horn
graph. A \textbf{Horn graph} is a labelled directed graph $G=(V,E,\varphi)$.
\begin{itemize}
\item $V$ is a union of two disjoint sets of vertices; Horn term
  vertices $V_T$ and disjunction vertices $V_\vee$. Each vertex $u \in
  V_\vee$ has eactly one outgoing edge to some $v \in V_T$.
\item $E$ is a set of labelled edges, which are expressed in the form
  $(u,v,\theta)$ where $\theta$ is a finite map from variables to
  variables.
\item $\varphi: V_T \rightarrow L$ is a labelling function for Horn
  term vertices, where L is a set of elements of a predicate variable
  in the form $P(x_1, \ldots, x_{arity(P)})$ or a linear inequality
  $a_1 x_1 + \cdots + a_n x_n \leq b$.
\end{itemize}
There are two kinds of edges in a Horn graph;
\begin{itemize}
\item for $u \in V_T, v \in V_\vee$, an edge takes a form
  $(u,v,\theta)$ written $u \xrightarrow{\theta} v$, or,
\item for $u \in V_\vee, v \in V_T$, an edge takes a form
  $(u,v,\emptyset)$ written $u \rightarrow v$, where the mapping is
  empty.
\end{itemize}

The meaning $\llbracket G \rrbracket $ of a Horn graph $G$ is a set of
Horn clauses encoded as follows.

\begin{align*}
\mathcal{T}_v & = \prod_{u \rightarrow v}
 \left\lbrace \theta (\varphi (t)) \mid t \xrightarrow{\theta} u \right\rbrace \\
\mathcal{HC}_{(t_1,\ldots,t_k),t_\star} & = \forall \left(\textsc{FV}(\mathbf{t}) \cup \textsc{FV}(t_\star) \right);
 \left\lbrace \left( \bigwedge t_i \right) \Longrightarrow t_\star \right\rbrace \\
\llbracket G \rrbracket & = \bigwedge_{v \in V_T}
 \left(\bigwedge_{\mathbf{t} \in \mathcal{T}_v} \mathcal{HC}_{\mathbf{t},\varphi (v)} \right)
\end{align*}

Without loss of generality, we limit linear expression labels on $V_T$
vertices with incoming edges to either $0 \leq -1$, meaning $\bot$, or
$0 \leq 0$, meaning $\top$.

@@

\subsection{Simple solution}

We consider below a Horn graph that is acyclic, and we consider the case
that all disjunction vertices has exactly one predecessor.

For simplicity of discussion, we define $k_v$ the effective arity of
the Horn clause $\mathcal{HC}_{\mathbf{t},\varphi(v)}$.
\begin{align*}
k_v =
\begin{cases}
arity(P) & \mbox{if } \varphi(v) = P(x_1,...,x_{arity(P)}) \\
0 & \mbox{if } \varphi(v) = \top \mbox { or } \varphi(v) = \bot
\end{cases}
\end{align*}
We let $K$ denote the maximum of $k_v$ over all $v$.

We define the relation $\leadsto$ over $V_T$ as follows.
\[ u \mathop{\leadsto}^\theta v \Longleftrightarrow
\exists a \in V_\vee; \left\lbrace (u,a,\theta), (a,v,\emptyset) \right\rbrace \subseteq E \]
We may omit $\theta$ and write $u \leadsto v$ for convenience in later
discussions.

By the assumption that $G$ is acyclic, $\leadsto^+$ is a
well-founded relation. We assume without loss of generality that $V_T$
contains the greatest element $v_\bot$ with respect to $\leadsto^+$;
otherwise we can always implicitly assume $\varphi(v_\bot) = \top$
such that $\forall v \in V_T \setminus \{v_\bot\}; v
\mathop{\leadsto}^\emptyset v_\bot$.

For each $v \in V_T$, we define
$\Delta(v) = (\mathbf{a}_v \mathbf{x}_v \leq b_v, C_v)$
, where
\begin{itemize}
\item $\mathbf{a}_v \mathbf{x}_v \leq b_v$ is a linear inequality that
  contains coefficient variables $\mathbf{a}_v$, and
\item $C_v$ is a set of linear expressions on coefficient variables.
\end{itemize}
Intuitively, the pair $(\mathbf{a} \mathbf{x} \leq b, C)$ denotes the
set of inequalities:
\begin{align*}
\left\lbrace
 \mathbf{a} \mathbf{x} \leq b \middle|
 \exists \left( \textsc{FV}(C)
  \setminus (\mathbf{a} \cup \left\lbrace b \right\rbrace
 \right); C
\right\rbrace
\end{align*}

$\Delta(v)$ is defined by well-founded induction on $v$ with respect
to the relation $\leadsto^+$. Then, $C_v$ in
$\Delta(v) = \left( \mathbf{a}_v \mathbf{x}_v \leq b_v, C_v \right)$
can be shown as follows.

\begin{align*}
C_v = & \bigcup_{u \leadsto v} C_u
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u \leadsto v \\
\left\lbrace
 \mathbf{a}_{v,i} = \sum_{u \mathop{\leadsto}^\theta v} \mathbf{a}_{u, \theta^{-1} (i)},
 b_v \geq \sum_{u \mathop{\leadsto}^\theta v} b_u
\right\rbrace \cup \\
\hspace{2cm} \bigcup_{k_v < i \leq K} \left\lbrace \mathbf{a}_{v,i} = 0 \right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}_v) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_v = \lambda_v \mathbf{a}_0,
 b_v = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x}_v \leq b_0
\end{cases}
\end{align*}

@@

\subsection{DAG problem}

@@

We provide $\textsc{Copy}_v$ procedure to rename all free variables
for avoiding name collision.
\[ \textsc{Copy}_v (x_V) = x_{V \cup \left\lbrace v \right\rbrace} \]
All variables with the same name are then distinguished by the label of copy origin
nodes $V$. For consistency, the variables without the label should be
interpreted as that it has $\emptyset$ for origin nodes.

We give a constraint function over a copy tree
$\hat \Delta(v,e)$ for $v \in V_C, e \in E$ as
\[ \hat \Delta (v,e) = \textsc{Copy}_v (\Delta_{V'} (\xi (v)))
\text{ such that } V' = \left\lbrace v' \middle| (v, v',e) \in E_S \right\rbrace \]
extending $\Delta$ function to
$\Delta_V (v) = (\mathbf{a}_{v,\emptyset} \mathbf{x}_v \leq b_{v,\emptyset}, C_v)$
as
\begin{align*}
C_v = & \bigcup_\circ C_u \cup \bigcup_\bullet C_u \cup
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u \rightarrow v \\
\left\lbrace
 \mathbf{a}_{v,\emptyset,i} =
  \sum_\circ \mathbf{a}_{u,U,\theta^{-1}(i)} +
  \sum_\bullet \mathbf{a}_{u,U, \theta^{-1}(i)}
\right\rbrace \cup \\
\hspace{2cm} \left\lbrace
 b_{v,\emptyset} \geq
  \sum_\circ b_{u,U} +
  \sum_\bullet b_{u,U}
\right\rbrace \cup \\
\hspace{2cm} \bigcup_{k_v < i \leq K} \left\lbrace \mathbf{a}_{v,\emptyset,i} = 0 \right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}_v) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_{v,\emptyset} = \lambda_v \mathbf{a}_0,
 b_{v,\emptyset} = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x}_v \leq b_0
\end{cases}
\end{align*}
under
\[
\circ = \substack{
  u \mathop{\leadsto}^\theta v \\
  \nexists u' \in V; \xi (u') = u \\
  \Delta_V (u) = (\mathbf{a}_{u,U} \mathbf{x}_u \leq b_{u,U}, C_u)}
\qquad \bullet = \substack{
  u \mathop{\leadsto}^\theta v \\
  \exists u' \in V; \xi (u') = u \\
  \hat \Delta (u', u \rightarrow v) = (\mathbf{a}_{u,U} \mathbf{x}_u \leq b_{u,U}, C_u)} \]

@@

\begin{align*}
 \rho = \left\lbrace
  \left( P, \bigwedge_U \sigma(\mathbf{a}_{v,U} \mathbf{x}_v \leq b_{v,U}) \right) \middle|
  \forall v \in V_T; \varphi(v) = P(\mathbf{x}_v)
 \right\rbrace
\end{align*}

@@

\subsection{Disjunctive problems}

We eliminate the restriction that all disjunction vertices $V_\vee$ have exactly one
predecessor. We define the universal set of selectors for disjunction vertices.
\[ \mathcal{S} = \prod_{v \in V_\vee} \left\lbrace v \mapsto u \middle| \exists \theta; (u,v,\theta) \in E \right\rbrace \]

We define a \textbf{selector set collection} \cite{albarghouthi13} for every Horn term vertices, which is a set of sets of selectors.

For $v \in V_T$, the selector set collection $\mathcal{X}_v = \left\lbrace S_{v,1}, \ldots, S_{v,\kappa_v} \right\rbrace$ consists of $\kappa_v$ sets of non-empty and non-overlapping subsets of $\mathcal{S}$. If $\nexists u; u \rightarrow v$, $\mathcal{X}_v = \emptyset$.

We arbitrarily choose the initial selector $s_0 \in \mathcal{S}$. We start from empty collections for all $v$.

\bibliographystyle{plain}
\bibliography{./biblio}
